---
output: 
  md_document:
    variant: gfm
---

<!-- Please edit README.Rmd - not README.md -->

# Week 08: Functional Programming

At this point you should be comfortable writing functions as part of normal
coding practice. This week's questions focus on some of the techniques and 
tools that are collectively known as *functional programming*. While all of
these questions can be completed using base R functions, these questions 
expect you to use functions from the `{purrr}` package, which have a 
more consistent API and are less likely to trip you up with subtle 
idiosyncrasies. 

```{r}
# `{purrr}` is loaded as part of the tidyverse:
suppressPackageStartupMessages(library(tidyverse))
```

# Prerequisites

*  These questions assume some familiarity with `for`-loops and lists in R. If 
   you're not familiar, please read the
   [lists chapter of R for Data Science](https://r4ds.had.co.nz/vectors.html#lists) 
   and the [loops chapter of Advanced R](https://adv-r.hadley.nz/control-flow.html#loops)
   before attempting these questions.

*  The [RStudio `{purrr}` cheatsheet](https://github.com/rstudio/cheatsheets/blob/main/purrr.pdf)
   is a useful resource to get quick summaries of how the most useful `{purrr}` 
   functions work. This is well worth skimming over before attempting these 
   questions - you'll likely find it a useful reference later on.
   
*  The [Advanced R chapter on functionals](https://adv-r.hadley.nz/functionals.html)
   gives a great overview of function programming with `{purrr}`. 

# Q1: For-loops and `map()`

This example will walk through the use of for-loops and hopefully motivate
the use of a very useful tool in R programming - the `map()`
function from the package `{purrr}`.

1.  The following code takes a vector of 5 random sentences, then loops over the
    vector to reverse the words in each sentence:
    
    ```{r}
    # This is the 'input'
    my_sentences <- sentences[1:5]
    
    # Set up an output vector
    reversed_sentences <- vector("list", 5)
    
    # Reverse each sentence and assign to the output vector
    for (i in seq_along(my_sentences)) {
      sentence     <- my_sentences[i]
      words        <- strsplit(sentence, split = " ")[[1]]
      rev_sentence <- paste(rev(words), collapse = " ")
    
      reversed_sentences[i] <- rev_sentence
    }
    
    reversed_sentences
    ```
    Create a function `str_reverse()` to do the job of reversing the
    sentence. This should take two arguments:
    * `x`: The string to reverse
    * `split`: The character used to split and recombine the string. This
      should have a default value of `" "`.
    
    Using this function, refactor the above code. The for-loop should now
    look like this:
    
    ```
    for (i in seq_along(my_sentences)) {
      reversed_sentences[i] <- str_reverse(my_sentences[i])
    }
    ```
    
2.  Now, create a function `my_map()` which abstracts the above logic - i.e.
    does the hard work of creating and executing a for-loop for you. This
    function should have the following arguments:
    * `x`: An object to iterate over
    * `f`: A function to be applied to each element of `x`
    
    Refactor your answer to part 2 to use `my_map()`. Your code, not including
    the definition for `my_map()`, should now look something like this, and
    should produce exactly the same output as part 2:
    ```
    reversed_sentences <- my_map(my_sentences, str_reverse)
    ```
   
3.  Notice that `str_reverse()` has the following behaviour: 
    ```
    str_reverse("hello world!", split = "")
    
    #> [1] "!dlrow olleh"
    ```
    Add a `...` argument to `my_map()`, giving arguments to be 
    'passed' to `f()`. Using this new functionality, try using `my_map()`
    and `str_reverse()` to reverse words rather than sentences - e.g.
    
    ```
    my_words <- c("one", "two", "three")
    
    # Changing the `sep` argument should mean that `str_reverse` internally
    # splits strings into vectors of characters rather than vectors of words
    my_map(my_words, str_reverse, sep = "")
    
    #> [[1]]
    #> [1] "eno"
    #> 
    #> [[2]]
    #> [1] "owt"
    #> 
    #> [[3]]
    #> [1] "eerht"
    ```

# Note: why `map()`?

The function just created mimics the behaviour of `purrr::map()`. You may wonder, 
why use `map()` if you can use a `for`-loop?

*  Once you've internalised the usage of `map()` (and other similar functions), 
   you'll find that 90% of the time it's much easier to write code in this way. 
   The only possible exception is when you need to recursively modify an object 
   which exists outside of your loop.
   
*  While `for`-loops can be performant if written correctly, there are a few
   gotchas that you need to be aware of to avoid bottlenecks. `map()` tends
   to be fast by default.
   
*  `map()` encourages a more 'functional' style of programming, which is worth
   leaning into to become a really effective R programmer. 
   
# 2. `map()`, `map2()`, `imap()` and `pmap()`

1.  Read the documentation for `purrr::map()` and use this function to simplify
    the following code:
    ```{r}
    sample_sizes <- c(3, 5, 7)
    samples      <- vector("list", length = 3)
    
    for (i in 1:3) {
      samples[[i]] <- rnorm(sample_sizes[i])
    }
    
    samples
    ```
 

2.  Read the documentation for `purrr::map2()` and use it to simplify the following 
    code:
    ```{r}
    sample_sizes  <- c(3, 5, 7)
    distributions <- list(rnorm, runif, rexp) 
    samples       <- vector("list", length = 3)
    
    for (i in 1:3) {
      samples[[i]] <- distributions[[i]](sample_sizes[[i]])
    }
    
    samples
    ```
    Hint: try writing a function to do the work of the code inside the `for`-loop
    `distributions[[i]](sample_sizes[[i]])`. This function should take two
    arguments: (1) the size of the sample to produce and (2) the function to use
    when producing the sample.
   
3.  Read the documentation for `imap()` and use it to simplify the following 
    code:
    ```{r}
    mtcars_list <- as.list(mtcars)[1:4]
    
    for (col in names(mtcars_list)) {
      mtcars_list[[col]] <- paste0(
        "Median `", col, "`: ", 
        median(mtcars_list[[col]])
      )  
    }
    
    mtcars_list
    ```
    
4.  Read the documentation for `purrr::pmap()`. Using this function, produce 
    3 histogram plots, one for each row of the following example dataset. 
    Each histogram should have a title that refrences the `id` and `distribution`
    columns.
    
    ```{r, fig.show="hold", out.width="33%"}
    # Make random numbers reproducible
    set.seed(100)
    
    # Dataset containing samples for 3 distributions
    plot_data <- tibble(
      id           = sample(1:100, 3),
      distribution = c("normal", "uniform", "exponential"),
      values       = list(rnorm(1000), runif(1000), rexp(1000))
    )
    ```

    Hint: You can create a quick histogram using `qplot()`, e.g. 
    `qplot(rnorm(1000))`
    
# 3. Anonymous functions
A key characteristic of most `{purrr}` functions is that they tend to take 
other functions as arguments. Creating a named, custom function every time you
use a tool like `map()` can feel cumbersome, so it's worth knowing how to 
create anonymous functions to make your code easier to write and read. These 
questions outline a few ways of doing this.

1.  The following code investigates the relationship between diamond size and 
    price by fitting a linear relationship between the two for each value of 
    `cut`, then extracting the 
    [coefficient of determination $R^2$](https://en.wikipedia.org/wiki/Coefficient_of_determination) 
    for each model.
    ```{r}
    # 1. Create a split version of `diamonds`
    diamonds_split <- split(diamonds, diamonds$cut)
    
    # 2. Define a function to compute R^2
    size_to_price_relationship_strength <- function(data) {
      data %>% 
        lm(price ~ x * y * z, data = .) %>% 
        summary() %>% 
        .$r.squared
    }
    
    # 3. 
    diamonds_split %>% 
      map(size_to_price_relationship_strength)
    ```
    Rewrite step #3 in the above code to use an anonymous function. This should
    be as simple as copy/pasting the definition for `size_to_price_relationship_strength()`
    into the `map()` call.
    
2.  Read the documentation for `purrr::map()` and repeat part 1 using
    formula syntax. Hint: you may also find the documentation for 
    `rlang::as_function()` helpful.
    
3.  The pipe `%>%` can also be used to create compact anonymous functions. 
    Read the examples in `` ?`%>%` `` and use repeat part 1 using this 
    syntax.
    
4.  Comment on the relative strengths/weaknesses of each technique:
    *  Creating named custom functions beforehand
    *  Defining anonymous function using R's default `function` syntax
    *  `{purrr}`'s formula syntax
    *  `{magrittr}`'s pipe syntax
    
# 4. Other iterators from `{purrr}`

`{purrr}` has several other functions which are less frequently needed but
are still worth knowing about.

1.  Read the documentation for `accumulate()`. Use this function to create 
    alternative implementations for `cumsum()` and `cumprod()`

2.  Read the documentation for `reduce()`. Use this function to create a more
    powerful version of `inner_join()`, `inner_join2()`, which should behave
    identically to `inner_join()`, but should allow the user to
    supply a list of dataframes as the `x` argument, in which case they will
    be recursively joined. 
    Hint: You'll need to think carefully about the default behaviour of
    the `suffix` argument.

Other functions from `{purrr}` are beyond the scope of these questions but
are still worth knowing about! These are a bit simpler and help you solve
common problems, so are worth knowing about! Some of the most useful are 
`pluck()`, `chuck()`, `walk()`, `compact()`, `compose()`, `set_names()`,
`keep()` and `flatten()`. 