---
title: "Week 4 Functions: Solutions"
author: "Stavrina Dimosthenous"
date: "18/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(rlang))
```

<!-- CSS Styles etc. -->
<style>
colourful {
    color:#E66100;
}
</style>
<!-- CSS Styles etc. -->

## Q 1
### Q 1 Function declaration
```{r, include=TRUE, echo=TRUE}
prop_na <- function(property_vec) {
  
  # This function takes in a list / vector / column and calculates the proportion
  # of NA values.
  
  sum(is.na(property_vec)) / length(property_vec)
} 

var_coeff <- function(property_vec) {
  
  # This function takes in a list / vector / column and calculates the coefficient
  # variation, which is the standard deviation divided by the mean and can be defined as
  # the extent of data point dispersion around the mean of a distribution.
  
  
  sd(property_vec,     na.rm = TRUE) / mean(property_vec,     na.rm = TRUE)
}

standardise_date <- function(dates, input_format, output_format = "%Y-%m-%d") {
  
  # This function parses a date passed in the input format to a standard output format
  # output format can be modified.
  
  dates <- strptime(dates, input_format)
  strftime(dates, output_format)
  
}
```
### Q 1 working area
#### Stats functions
```{r, include=TRUE, echo=TRUE}

storms %>% 
  group_by(status) %>% 
  summarise(
    
    
    # Proportions of `NA` values for ts_diameter and hu_diameter
    across(c(ts_diameter, hu_diameter), prop_na),
    
    # Coefficients of variation for wind and pressure
    across(c(wind, pressure), var_coeff),

    .groups = "drop"

  )

```
#### Dates functions
```{r, include=TRUE, echo=TRUE}

# Some less-than-ideally formatted dates
dates1 <- c("20220103", "20220415", "20220418")
dates2 <- c("Mon, May 02 2022", "Thu, Jun 02 2022", "Fri, Jun 03 2022")
dates3 <- c("29 August 22", "26 December 22", "27 December 22")

# -- From here on is the code to simplify ------------------------

(dates1_formatted <- standardise_date(dates1, "%Y%m%d", "%d %B %y"))

(dates2_formatted <- standardise_date(dates2, "%a, %b %d %Y", "%d %B %y"))

```

## Q 2
### Q 2 Function declaration area
```{r, include=TRUE, echo=TRUE}
#### Info print
print_info <- function (.data, quiet = FALSE, ...) {
  
  # This function takes in a data frame and prints the number of rows, columns
  # and total cells within the data frame.
  
  if (!quiet) {
    cat(
      ...,
      paste("Rows:       ", nrow(.data)),
      paste("Columns:    ", ncol(.data)),
      paste("Total cells:", nrow(.data) * ncol(.data)),
      sep = "\n"
      )
    return(.data)
  }
  invisible(NULL)
}


#### Pipable gsub() wrapper
gsub2 <- function(x, pattern, replacement) {
  # This function is a gsub wrapper that works with the pipe
  x <- gsub(pattern, replacement, x)
  x
}
```
### Q 2 working area
#### Piping
```{r, include=TRUE, echo=TRUE}
starwars %>%
  print_info() %>%
  count(homeworld, species) %>%
  print_info()
```

```{r, include=TRUE, echo=TRUE}
c("string 1", "string 2") %>% gsub2("ing", "")

#### Part 3
c("string 1", "string 2") %>% gsub("ing", "", .)
```

## Q 3
### Q 3 Function declaration area
N.B. The function `either` and the infix function `%or%` are identical, I tried two different ways but the function body could be a verbatim copy.
```{r, include=TRUE, echo=TRUE}


# Did not change the name here but the function could be renamed to select_nonempty
either <- function(x, y) {
  # This function will return x if it is not empty, i.e. not NULL or an empty
  # list or vector.
  if (length(x) > 0) x else y
}

`%or%` <- function(x, y) {
  # This function is identical to the above function, either.
  # This function will return x if it is not empty, i.e. not NULL or an empty
  # list or vector.
  if (length(x) > 0) x else y
}

`%splice%` <- function(x, y) {
  # This function takes in two lists or vectors of the same length and returns a
  # list that places their identical indices next to each other.
  # Will not work for strings or characters.
  if (length(x) == length(y)) {
    c(rbind(x, y))
  }
  else {
    stop("Please enter two objects of equal length.")
  }
}


##### Ignore this will attempt some other time
`%|e|%` <- function(x, y) {
  tryCatch()
}

```
### <a name="work-area-3"></a>Q 3 Working area
```{r, include=TRUE, echo=TRUE, results="hide"}

either(1, 2)

either(NULL, 2)

either(NA, 2)

either(c(), 2)

c("") %or% 2

NULL %or% 2

NA %or% 2

"" %or% 2

1 %||% 2

c("") %||% 2

NULL %||% 2

NA %||% 2

"" %||% 2

```
#### Compare `%or%` and `%||%`

`%or%` takes two objects and returns the first object if it is of length greater than 0, or the second object if it is empty. Likewise, `%||%` takes two objects, however, it will only return the second object if the first object evaluates to `NULL`, otherwise it will return empty objects passed as the first object. A few examples and what they evaluate to can be found in the [above](#work-area-3) working area.

#### Splicing
```{r, include=TRUE, echo=TRUE}

1:3 %splice% 11:13

"kitten" %splice% "dog"

```
#### Bonus: %|e|%
N.B. Yet to be attempted
```{r, include=TRUE, echo=TRUE}
```

## <colourful>Q 4</colourful>
### Q 4 Function declaration area
```{r, include=TRUE, echo=TRUE}

# Could be renamed to copy_as_text

as_text <- function (df, na = "", headers = TRUE, ...) {
  # 1. Looping over each column in the data, transform the column
  #    to character format, then replace any `"NA"` values with `""`
  
  for (col_name in colnames(df)) {
    col_formatted   <- format(df[[col_name]], ...)
    col_na_replaced <-
      ifelse(col_formatted == "NA", na, col_formatted)
    
    df[[col_name]] <- col_na_replaced
    
  }
  
  if (headers == FALSE) {
    colnames(df) <- NULL
  }
  
  # 2. Create text representations of the table body and headers. `"\t"`
  #    means 'new cell' and `"\r\n"` means 'new row'. Don't worry about
  #    `do.call()` - this might be covered in a future week.
  body_text <- do.call(paste, args = c(df, sep = "\t", collapse = "\r\n"))
  headers_text <- paste(colnames(df), collapse = "\t")
  
  # 3. Combine headers and body (and return)
  paste(c(headers_text, body_text), collapse = "\r\n")
  
}

# Could be renamed to copy_to_clipboard
to_clipboard <- function (df, na = "", headers = TRUE, ...) {
  # This function takes in a data frame and copies it to the clipboard as text

  # Convert data frame to text
  my_df <- as_text(df, na, headers, ...)

  # Write text to clipboard
  writeClipboard(my_df)
  
  
  message(cat("Your", nrow(df) ,"x",  ncol(df),"data frame has been copied to your clipboard."))
  
  # Return data frame
  
  invisible(df)
}
```

## Q 5
### Q 5 Function declaration area
```{r, include=TRUE, echo=TRUE}

#### Aggregate function ####
aggregate <- function (.data, ..., .cols, .fn = sum) {
  .data %>% group_by(...) %>% 
    summarise(
      across(c({{ .cols }}), .fn),
      .groups = "drop"
  )
}

#### Disaggregate function ####

disaggregate <- function (.data, ..., .cols, .fn = sum) {
  .data %>% 
    group_by(across(!c(..., {{ .cols }}))) %>%
    summarise(
      across(c({{ .cols }}), .fn),
      .groups = "drop"
      )
}

#### Bar Plot Function ####

bar_plot <- function (.data, x, y, fill = NULL, .fn = mean, facet = NULL, position = "dodge") {
  df <- aggregate(.data, {{ y }}, {{ fill }}, .cols = {{ x }}, .fn = {{ .fn }})
  
  df %>% ggplot(mapping = aes({{ x }}, {{ y }}, fill = {{ fill }})) +
    facet_wrap(vars({{ facet }})) +
    geom_col(position = position)
}

```

### Q 5 Code area
```{r, include=TRUE, echo=TRUE}
#### Test bar plot ####
diamonds %>% bar_plot(price, cut, color)
```

## Q 6
### Q 6 Function declaration area
```{r, include=TRUE, echo=TRUE}
#### File search function ####

search_up <- function (file, dir = getwd()) {
  # This function will search for a file within the working directory. If the
  # file is not found then the function will go up a level. The function will
  # stop if the file is not found in the uppermost level. 
  # This function does not crawl through all directories.
  search_dir <- dir
  level <- dirname(search_dir)
  if (file %in% list.files(search_dir)) {

    search_dir
  }
  
  else if (level != search_dir) {
    
    search_up(file, dir = dirname(search_dir))
    
  }
  
  else {
    message("File not found")
    return(invisible(NULL))
    }
}

```

### Q 6 Code are
```{r, include=TRUE, echo=TRUE}

search_up("fake_file.txt")

```