---
output: md_document
---

<!-- Please edit README.Rmd - not README.md -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 4,
  fig.height = 4,
  fig.path = "README_files/figure-markdown_strict/"
)
```

# Week 05: Visualisation

In week 1 we briefly touched on some data visualisation using the `{ggplot2}` 
package. This problem sheet will provide a deeper dive into plotting data with
R, and hopefully bring you to a point where you're able to integrate visualisation
into your day-to-day analysis with R.

# Resources

You should refer to the following resources when attempting the problems:

*  [R for Data Science](https://r4ds.had.co.nz/data-visualisation.html) chapters
   3 and 28 introduce the ggplot2 API and 'grammar of graphics'

*  The [ggplot2 documentation](https://ggplot2.tidyverse.org/) is a good place
   to read about how individual functions work. There are also a number of 
   vignettes (called 'articles' on the website) which cover individual topics
   and concepts.
   
*  The [ggplot2 book](https://ggplot2-book.org/) is an excellent and fairly
   comprehensive resource. If you want to know how to do something, check the 
   book before stackoverflow :)
   
# Exercises

**Note**: Please make sure you complete all these questions using `{ggplot2}`.
R has other plotting functions available (e.g. `plot()`), but these exercises
are intended to introduce the 'grammar of graphics' approach. While other 
functions may get the job done, you'll gain much more valuable skills if you
stick to `{ggplot2}` for these problems.

## Q1. Plotting `txhousing`

1.  Create a subset of `txhousing` which only includes counties (hint: try
    looking at all the distinct values of `city` first).

2.  Create a plot which shows how the median house price for each county
    changes over time. You should use `geom_line()` to do this, and 
    counties should be differentiated using colour.
    
3.  Now add lines for every city in the `txhousing` dataset to your plot.
    These should be coloured light grey and should not be included in the
    colour legend. 
    
    Hints: 
    
    *  In your layer which includes every city, you'll need to specify
       that each line should be grey *outside* of any calls to `aes()`
       
    *  You'll also need to specify the `group` aesthetic so cities are
       plotted as individual lines
       
4.  Now, make your plot look nice by:
    *  Using a colourblind-friendly colour palette
    
    *  Using nice labels on the y-axis (see `scales::label_dollar()`)
    
    *  Changing the `x`, `y`, `colour` and `title` labels using `labs()`.
       Hint: If it's self-evident, you can remove a label by setting it
       to `NULL`.
       
    *  Applying a nicer theme than the default `theme_grey()`
    
## Q2. Why 'tidy' data?

This question will demonstrate how `{ggplot2}` is designed to work with 'tidy' 
(i.e. 'long') data. If these are unfamiliar concepts, please read 
[R for Data Science Ch12](https://r4ds.had.co.nz/tidy-data.html#case-study)
before attempting these problems.

1.  Create a version of the `economics`
    dataset where `pce`, `pop`, `psavert`, `uempmed` and `unemploy` are rescaled
    to be between 0 and 1. You can use this function to do the rescaling:
    ```{r}
    rescale01 <- function(x) {
      rng <- range(x, na.rm = TRUE)
      (x - rng[1]) / (rng[2] - rng[1])
    }
    ```

2.  Now, create a plot of your rescaled dataset where the rescaled columns
    are all represented as different coloured lines, which should all be
    shown on the same panel.
    
3.  In part 2 you should have needed 5 separate calls to `geom_line()`. This
    part will demonstrate how you can simplify your plot code by reformatting
    your data. 
    
    Create a new version of the `economics` dataset in the following steps:
    
    *  Use `pivot_longer()` to pivot the columns `pce`, `pop`, `psavert`, 
       `uempmed` and `unemploy`. Column names should end up in a single 
       column `variable` and values should end up in a single column
       `value`
       
    *  Use `group_by()`, `mutate()` and `rescale01()` to create a rescaled
       version of the `value` column. Call this new column `value01`. Don't
       forget to `ungroup()`!
       
    *  The above steps should be done in a single call using the pipe `%>%`.
    
4.  Using your dataset from part 3, recreate the plot from part 2. 
    
## Q3: Changing plot appearance

1.  Using your final plot from the previous question. Most of these can be completed
    using the `theme()` and `labs()` functions:
    *   Move the colour legend to the bottom of the plot
    *   Choose appropriate labels for the title, subtitle, legend and axes
    *   Increase the size of the title and centre the title text above the plot
    *   Remove the axis and 'ticks' for the y-axis
    *   Use a colourblind-friendly palette for the colour scale, and make the lines
        a bit thicker than the default
    *   Make the background of the whole plot`lightblue`
    
2.  The axis labels on the following plot are not very clear. Look at 
    [this post](https://stackoverflow.com/questions/41568411/how-to-maintain-size-of-ggplot-with-long-labels/66169251#66169251) on stackoverflow and find a good solution to the problem:
    ```{r}
    suppressPackageStartupMessages(library(tidyverse))
    
    mpg %>% 
      ggplot(aes(manufacturer)) +
      geom_bar()
    ```
    
    
## Q4: Using extensions

While `{ggplot2}` is a very powerful package, it doesn't do everything. 
Fortunately, many extensions exist, meaning the functionality you need is
probably out there somewhere. A list of some of the most popular extensions
can be found [here](https://exts.ggplot2.tidyverse.org/gallery/).

1.  Use `{geomtextpath}` to add labels directly to the lines in the following 
    plot. Remove the colour legend too as this will no longer be needed:
    ```{r}
    ggplot(diamonds, aes(depth, colour = cut)) +
      geom_density() +
      xlim(55, 70)
    ```
    
2.  Use `{ggpointdensity}` to improve the following plot:
    ```{r}
    txhousing %>% 
      ggplot(aes(log(sales), log(listings))) +
      geom_point()
    ```

3.  Use `{ggthemes}` to change the appearances of the previous two plots. Try
    to find a theme that's reasonable nice :) 

4.  Use `{patchwork}` to combine the previous two plots into a 
    single image.
    
## Q5. Some `{ggplot2}` subtleties

1.  We want a plot of the `diamonds` dataset showing `carat`/`price`. We want the
    points to be blue, but the following code isn't working quite right:
    ```{r}
    ggplot(diamonds, aes(carat, price, colour = "blue")) +
      geom_point()
    ```
    
    Try the following three methods of fixing it:
    *  Putting `colour = "blue"` in a different place
    *  Adding a (appropriately configured) call to `scale_colour_manual()`
    *  Adding a call to `scale_colour_identity()`
    
    What use-cases do you think each method would be most suited to?
    
2.  The `group` aesthetic is usually not needed, but is good to be aware of.
    Try creating a line-plot of the following data. You should get an error -
    try to fix this using the `group` aesthetic. Why is this needed?
    ```{r}
    df <- tibble(x = c("1", "2", "3", "4", "5"), y = 1:5)
    ```
    
3.  Re-create the plot from part 2 *without* using the `group` aesthetic. Do
    this by first applying an appropriate `mutate()` to the `x` variable in
    `df`.
    
4.  Why doesn't the following code work? Find 3 ways to fix it. 
    ```{r, eval = FALSE}
    ggplot(mpg, aes(displ, manufacturer, colour = cyl)) +
      geom_jitter() +
      scale_colour_viridis_d()
    ```
    (Hint: try adjusting the scale, adjusting the `colour` aesthetic and adjusting
    the data before plotting).
    
5.  The following plot is not very space-efficient. Create a version of the
    plot with no empty panels. 
    ```{r}
    ggplot(mpg, aes(y = fl, fill = factor(year))) +
      geom_bar() +
      facet_grid(cyl ~ class) +
      theme(legend.position = "bottom")
    ```
    